Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated

### 2024-10-23 submission/R2/manuscript-Academ-7C8RB54.qmd
clean manuscript

Beyond stimulus- and individual-related characteristics, evidence
suggests that the interaction between the two (i.e., the subjective
reaction associated with the experience of a given stimulus),
contributes to simulation monitoring decisions. For instance, the
intensity of experienced emotions have been shown to increase one's
sense of presence - the extent to which one feels like "being there", as
if the object of experience was real - when engaged in a fictional movie
or a VR environment [@sanchez2005presence; @makowski2017being]. Indeed,
participants' self-reported emotional arousal were found to
significantly predict the probability that they would perceive images as
real [@azevedo2020body]. Conversely, beliefs that emotional stimuli were
fake (e.g., that emotional scenes were not authentic but instead
involved actors and movie makeup) were found to result in emotion
down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In
line with these findings, studies on susceptibility to fake news have
also found heightened stimulus emotionality to be associated with
greater belief [@martel2020reliance; @bago2022emotion], and higher
neurophysiological arousal was predictive of judging realistic images as
real [@azevedo2020body]. Additionally, other factors, such as the
stimuli's perceived self-relevance [@sperduti2016paradox;
@goldstein2009pleasure], as well as familiarity [@begg1992dissociation],
could also play a role in guiding our appraisal of a stimulus.**For
instance,** @miller2023 **reported that participants were more likely to
mistakenly identify AI-generated faces as real because they perceived
them as more familiar.**

@@

**Procedure.** In the first part of the study, participants answered a
series of personality questionnaires presented in the order below. These
include the *Mini-IPIP6* <!-- (24 items) [@sibley2011]  --> [24 items,
@sibley2011] measuring 6 personality traits, the *SIAS-6* and the
*SPS-6* <!-- (6 items each) [@peters2012development] --> [6 items each,
@peters2012development] assessing social anxiety levels, 5 items we
devised pertaining to expectations about AI-generated image technology
("I think current Artificial Intelligence algorithms can generate very
realistic images"), **to potentially test and mitigate the potential
effect of expectations/beliefs about AI.** **These items were** mixed
with 5 items from the general attitudes towards AI scale to lower the
former's saliency and the possibility of it priming the subjects about
the task, <!-- (*GAAIS*) [@schepman2020initial]. --> [*GAAIS*,
@schepman2020initial] the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] --> [30 items,
@jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] --> [18 items,
@freeman2021revised] measuring 2 dimensions related to paranoid
thinking; and the *IUS-12* <!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing] measuring intolerance to uncertainty.
Self-rated attractiveness was also assessed using 2 items - one
measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->
["How attractive would you say you are?", @marcinkowska2021self] and the
other measuring physical attractiveness
<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the
average", @spielmann2020predictive]. 3 attention check questions were
also embedded in the surveys. **All Cronbach's alpha values were within
the acceptable to excellent range, except for the neuroticism subscale
of the Mini-IPIP6 and the negative subscale of the GAAI, which were
poor, and the Expectations about AI scale, which was questionable (Gliem
& Gliem, 2003; see supplementary material for the details of the
reliability analysis).**

In the second part of this study, images of neutral-expression faces
from the validated American Multiracial Face Database
<!-- (AMFD) [@chen2021broadening]  --> [AMFD, @chen2021broadening] were
presented to the participants for 500ms each, in a randomized order,
following a fixation cross display (750 ms). T**he decision to present
the faces for 500 ms was based on pilot studies, which demonstrated that
this duration provides a sufficient perceptual window for
decision-making and aligns with previous research indicating stable
judgment levels and increased confidence beyond this exposure time
[@willis2006first].**

>>>>

Beyond stimulus- and individual-related characteristics, evidence
suggests that the interaction between the two (i.e., the subjective
reaction associated with the experience of a given stimulus),
contributes to simulation monitoring decisions. For instance, the
intensity of experienced emotions have been shown to increase one's
sense of presence - the extent to which one feels like "being there", as
if the object of experience was real - when engaged in a fictional movie
or a VR environment [@sanchez2005presence; @makowski2017being]. Indeed,
participants' self-reported emotional arousal were found to
significantly predict the probability that they would perceive images as
real [@azevedo2020body]. Conversely, beliefs that emotional stimuli were
fake (e.g., that emotional scenes were not authentic but instead
involved actors and movie makeup) were found to result in emotion
down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In
line with these findings, studies on susceptibility to fake news have
also found heightened stimulus emotionality to be associated with
greater belief [@martel2020reliance; @bago2022emotion], and higher
neurophysiological arousal was predictive of judging realistic images as
real [@azevedo2020body]. Additionally, other factors, such as the
stimuli's perceived self-relevance [@sperduti2016paradox;
@goldstein2009pleasure], as well as familiarity [@begg1992dissociation],
could also play a role in guiding our appraisal of a stimulus. **For
instance,** @miller2023 **reported that participants were more likely to
mistakenly identify AI-generated faces as real because they perceived
them as more familiar.**

@@

**Procedure.** In the first part of the study, participants answered a
series of personality questionnaires presented in the order below. These
include the *Mini-IPIP6* <!-- (24 items) [@sibley2011]  --> [24 items,
@sibley2011] measuring 6 personality traits, the *SIAS-6* and the
*SPS-6* <!-- (6 items each) [@peters2012development] --> [6 items each,
@peters2012development] assessing social anxiety levels, 5 items we
devised pertaining to expectations about AI-generated image technology
("I think current Artificial Intelligence algorithms can generate very
realistic images"), **to potentially test and mitigate the potential
effect of expectations/beliefs about AI.** **These items were** mixed
with 5 items from the general attitudes towards AI scale to lower the
former's saliency and the possibility of it priming the subjects about
the task, <!-- (*GAAIS*) [@schepman2020initial]. --> [*GAAIS*,
@schepman2020initial] the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] --> [30 items,
@jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] --> [18 items,
@freeman2021revised] measuring 2 dimensions related to paranoid
thinking; and the *IUS-12* <!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing] measuring intolerance to uncertainty.
Self-rated attractiveness was also assessed using 2 items - one
measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->["How attractive would you say you are?", @marcinkowska2021self] and the
other measuring physical attractiveness
<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the
average", @spielmann2020predictive]. 3 attention check questions were
also embedded in the surveys. **All Cronbach's alpha values were within
the acceptable to excellent range, except for the neuroticism subscale
of the Mini-IPIP6 and the negative subscale of the GAAI, which were
poor, and the Expectations about AI scale, which was questionable (Gliem
& Gliem, 2003; see supplementary material for the details of the
reliability analysis).**

In the second part of this study, images of neutral-expression faces
from the validated American Multiracial Face Database <!-- (AMFD) [@chen2021broadening]  -->[AMFD, @chen2021broadening] were
presented to the participants for 500ms each, in a randomized order,
following a fixation cross display (750 ms). T**he decision to present
the faces for 500 ms was based on pilot studies, which demonstrated that
this duration provides a sufficient perceptual window for
decision-making and aligns with previous research indicating stable
judgment levels and increased confidence beyond this exposure time
[@willis2006first].**
